
---
title: "STA314 Tutorial 1"
subtitle: "R Review and Linear Regression"
author: "Teaching Team of STA314"
date: "Fall 2026"
output: pdf_document
---

# Quick R Review

## Load and inspect the data

We will use the `Credit` dataset from the `ISLR2` package.

```{r load-data}
library(ISLR2)
data("Credit")
```

```{r inspect-data}
# First six observations
head(Credit)

# Variable names
names(Credit)

# Structure and variable types
str(Credit)

# Number of observations and variables
dim(Credit)
```

**Quick check:** What information does `str(Credit)` provide that
`head(Credit)` does not?

## Access variables and calculate summaries

The `$` operator extracts one variable from a data frame.

```{r access-variables}
# Extract Income
Credit$Income

# First five Income values
Credit$Income[1:5]

# Summary statistics
mean(Credit$Income)
mean(Credit$Balance)
max(Credit$Income)
```

Before running the following commands, predict what each will return.

```{r predict-output}
Credit$Balance[1:3]
mean(Credit$Rating)
```

## Logical conditions and subsetting

```{r subsetting}
# Logical condition
Credit$Student == "Yes"

# Select all student rows
Credit[Credit$Student == "Yes", ]

# Select particular rows and columns
Credit[
  Credit$Student == "Yes",
  c("Income", "Balance")
]

# Examine categorical variables
levels(Credit$Student)
levels(Credit$Region)
table(Credit$Region)
```

Recall that `data[rows, columns]` specifies the observations and variables to
retain.

# Guided Example: Credit Card Data

The `Credit` dataset contains simulated information on 400 individuals. We
will use `Balance` as the response and consider predictors including `Income`,
`Rating`, `Student`, and `Region`.

## Explore Income and Balance

```{r scatterplot}
plot(
  Credit$Income,
  Credit$Balance,
  xlab = "Income",
  ylab = "Balance",
  pch = 19
)
```

Based only on this plot:

1. Does there appear to be a relationship between Income and Balance?
2. Does the relationship appear approximately linear?
3. Is there substantial variability around the relationship?

## Fit a simple linear regression

Consider

$$
\operatorname{Balance}_i
= \beta_0+\beta_1\operatorname{Income}_i+\epsilon_i.
$$

```{r simple-regression}
fit_slr <- lm(
  Balance ~ Income,
  data = Credit
)

summary(fit_slr)
coef(fit_slr)
```

The fitted model is approximately

$$
\widehat{\operatorname{Balance}}
= 246.51+6.05(\operatorname{Income}).
$$

Because Income is recorded in thousands of dollars, the estimated slope means
that an additional $\$1{,}000$ of Income is associated with an estimated 6.05
unit increase in Balance, on average.

### Add the fitted line

```{r fitted-line}
plot(
  Credit$Income,
  Credit$Balance,
  xlab = "Income",
  ylab = "Balance",
  pch = 19
)

abline(
  fit_slr,
  col = "red",
  lwd = 2
)
```

### Examine the residuals

```{r residual-plot}
plot(
  fitted(fit_slr),
  residuals(fit_slr),
  xlab = "Fitted values",
  ylab = "Residuals",
  pch = 19
)

abline(
  h = 0,
  col = "red",
  lty = 2,
  lwd = 2
)
```

What pattern would we hope to observe if the linear model were appropriate?
Do the residuals appear completely random?

## Predictions and intervals

Suppose we want predictions for individuals with Income equal to 20, 40, and
80.

```{r new-income-data}
new_data <- data.frame(
  Income = c(20, 40, 80)
)

new_data
```

### Point predictions

```{r point-predictions}
predict(
  fit_slr,
  newdata = new_data
)
```

### Confidence intervals for the mean response

```{r confidence-intervals}
predict(
  fit_slr,
  newdata = new_data,
  interval = "confidence",
  level = 0.95
)
```

### Prediction intervals for new individuals

```{r prediction-intervals}
predict(
  fit_slr,
  newdata = new_data,
  interval = "prediction",
  level = 0.95
)
```

The prediction intervals are wider because they account for uncertainty about
the mean response and individual variability around the regression line.

## Multiple linear regression

Now model Balance using Income, Rating, and Student status.

```{r multiple-regression}
fit_mlr <- lm(
  Balance ~ Income + Rating + Student,
  data = Credit
)

summary(fit_mlr)
```

Answer the following:

1. Interpret the estimated Income coefficient while holding Rating and Student
   status fixed.
2. Interpret `StudentYes` while holding Income and Rating fixed.
3. Is `StudentYes` statistically significant at the 5% level?

## Region as a categorical predictor

The `Region` variable has three levels.

```{r examine-region}
levels(Credit$Region)
table(Credit$Region)
```

Explicitly set East as the reference category and fit a model using Region.

```{r region-model}
Credit$Region <- relevel(
  Credit$Region,
  ref = "East"
)

fit_region <- lm(
  Balance ~ Region,
  data = Credit
)

summary(fit_region)
```

Interpretation:

- The intercept is the estimated mean Balance among individuals in the East.
- `RegionSouth` is the estimated difference in mean Balance between the South
  and East.
- `RegionWest` is the estimated difference in mean Balance between the West
  and East.

The p-values test whether each regional difference is zero. They do not test
whether all three regional means are equal simultaneously.

### Examine the model matrix

```{r region-model-matrix}
head(
  model.matrix(fit_region)
)

contrasts(Credit$Region)
```

Although Region has three categories, R creates only two dummy variables
because East is the reference category.

## Interaction between Income and Student status

The additive model assumes the same Income slope for students and
non-students. To allow different slopes, fit an interaction model.

```{r interaction-model}
fit_interaction <- lm(
  Balance ~ Income * Student,
  data = Credit
)

summary(fit_interaction)
```

In an R formula,

```r
Income * Student
```

is shorthand for

```r
Income + Student + Income:Student
```

The Income slope for non-students is:

```{r nonstudent-slope}
coef(fit_interaction)["Income"]
```

The Income slope for students is:

```{r student-slope}
coef(fit_interaction)["Income"] +
  coef(fit_interaction)["Income:StudentYes"]
```

The p-value for `Income:StudentYes` tests whether the Income slopes differ
between students and non-students.

# Practice Exercise

## 1. Interpretation

Use the following model:

```{r exercise-a}
fit_A <- lm(
  Balance ~ Income + Rating + Student,
  data = Credit
)

summary(fit_A)
```

1. What is the estimated coefficient for `StudentYes`?
2. Interpret it in one sentence.
3. Is it statistically significant at the 5% level?

## 2. Interaction

```{r exercise-b}
fit_B <- lm(
  Balance ~ Income * Student,
  data = Credit
)

summary(fit_B)
```

1. What is the Income slope for non-students?
2. What is the Income slope for students?
3. Is there evidence that the slopes differ?

## 3. Prediction

Use the multiple regression model to predict Balance for two individuals:

- Person 1: Income = 40, Rating = 300, Student = No;
- Person 2: Income = 40, Rating = 300, Student = Yes.

```{r exercise-c}
new_people <- data.frame(
  Income = c(40, 40),
  Rating = c(300, 300),
  Student = c("No", "Yes")
)

predict(
  fit_A,
  newdata = new_people,
  interval = "confidence",
  level = 0.95
)
```

Why do the two individuals have different predicted Balance values even though
they have the same Income and Rating?
